Predicting Kamala Harris's Victory in the 2024 US Election

### 2024-11-01 paper/paper.qmd
Modify Prose over the paper

The 2024 U.S. Presidential Election is set to unfold in a climate of heightened political polarization, economic challenges, and demographic shifts, making it a pivotal event in shaping America's future. Predicting the outcome of this election is crucial for gaining insights into potential changes in political leadership and policy directions. Polling data has traditionally served as a primary tool for gauging voter sentiment; however, recent election cycles have exposed limitations in polling accuracy due to biases, non-response issues, and variations in poll quality. To address these challenges, this study proposes a model that incorporates these factors to enhance the reliability of election forecasts.

This paper develops a prediction model aimed at forecasting the electoral outcomes between Kamala Harris and Donald Trump in the 2024 election. Using a multiple linear regression framework, we analyze aggregated polling data across U.S. states, incorporating factors such as poll quality, transparency, sample size, geographic indicators, and poll timing. By integrating these variables, our model seeks to provide a nuanced forecast that accounts for regional differences and methodological variations across polls, offering a comprehensive view of the projected support for each candidate.

Our findings suggest a competitive election landscape, with Harris showing strong prospects in key battleground states while Trump retains support in traditional Republican strongholds. These projections underscore the enduring influence of geographic voting patterns and the impact of polling methodology on electoral forecasting. The study provides valuable insights for political analysts, campaign strategists, and the public, contributing to a better understanding of how polling data can be utilized to anticipate electoral outcomes.

The remainder of this paper is organized as follows: @sec-data describes the dataset and the preprocessing steps involved in preparing data for the model. @sec-model explains the modeling approach, including the selection of predictors and the structure of the regression model. @sec-results presents the primary findings, highlighting state-by-state predictions and overall electoral projections. Section @sec-discussion discusses the implications, limitations, and future directions for this analysis. The Appendix section provides supplementary information on methodology, diagnostics, technical details of the model setup, acknowledgments, and references.

@@

The dataset used in this analysis originates from @fivethirty, a trusted source known for its rigorous methodology in collecting and aggregating polling data to represent public opinion accurately. To maintain data integrity and ensure a high standard of reliability, only polls meeting stringent quality criteria are included in the dataset. Specifically, each poll must disclose crucial information, including the pollster’s name, the exact dates of the survey, the sample size, and the full methodological details. These details encompass the polling medium (e.g., phone, online), use of verified voter files, demographic weighting criteria, and any adjustments applied to reflect a representative sample.

@@

Once deemed credible, polls are incorporated into the FiveThirtyEight database. These polls then contribute to polling averages, forecast models, and political coverage, which depend heavily on the quality and reliability of public opinion data. This meticulous selection process ensures that the dataset captures a comprehensive and nuanced view of public sentiment, providing an accurate reflection of behavioral trends and preferences over time.

@@

The distribution reveals that most observations cluster around a support percentage of approximately 48%, indicating a substantial level of backing from the electorate. Additionally, a smaller proportion of polls show support exceeding 55%, suggesting that while Trump has a core base, many voters remain either indifferent or opposed to him. This visualization provides insight into both the concentration and variability of support within the surveyed population.

@@

The sample size indicates the number of respondents in each poll. The distribution in @fig-sample-size exhibits a right-skewed shape, suggesting that there are more observations with smaller sample sizes compared to larger ones. The peak of the distribution is around 1,000, indicating that this is the most common sample size used in the polls. Overall, the sample size data highlights that each poll contains a sufficient number of respondents to provide reliable insights.

@@

As shown in the @fig-end-date, the support percentage for Donald Trump fluctuated over time, with a slight upward trend as the election approached, indicated by the dashed trend line. This suggests that public opinion varied throughout the period, and Trump’s support percentage experienced an increase as election day drew closer. However, it still remained below 50%. This trend highlights the importance of monitoring real-time polling data, as sentiment may shift in the critical months leading up to a closely contested election.

@@

The table below (@tbl-prediction) summarizes the predicted results, showing Trump’s predicted percentage, the number of electoral votes in each state, and the predicted winner (either Trump or Harris). For instance:
- Alabama: Trump is predicted to win 53% of the vote, securing all 9 electoral votes.
- California: Trump is predicted to receive 34.39% of the vote, resulting in a victory for Harris, who takes California's 55 electoral votes.
- Florida: The model predicts a close race, with Trump at 48.89% of the vote, resulting in a Harris win in this critical battleground state.

@@

## Predicted Electoral Outcomes by State
The table in the [Appendix -@tbl-missing-states] serves as a supplement to the predicted electoral outcomes by state, filling in the gaps for states where the model couldn’t provide predictions. It allows for a complete view of electoral projections by assigning a likely outcome in each missing state based on historical data. This approach ensures every state is accounted for in the final electoral vote totals.

@@

The analysis reveals that state-specific predictors, such as historical voting patterns and days until election day, are significant influencers of each candidate’s polling outcomes. Traditional Democratic strongholds like California and New York strongly favor Harris, while Republican-dominated states such as Alabama and Wyoming align with Trump. This distribution reflects enduring geographic voting patterns and underscores the role of regional party loyalty.

@@

The model also highlights the critical role of battleground states, with states such as Florida and Arizona predicted to lean toward Harris. The influence of these states, where margins are often narrow, emphasizes the importance of campaign efforts and voter mobilization in these regions. The projections illustrate that minor shifts in voter sentiment within these states could decisively impact the overall election outcome because of the high number of electoral votes.

@@

Due to limited data available for predicting outcomes in every state, some state projections rely on historical voting patterns, with the assumption that these trends will persist. However, this approach may be challenged by the rapidly changing political landscape in the U.S., where evolving demographics and shifting public sentiment could lead to unexpected results.

@@

Non-response in survey research is critical to address because it introduces potential bias, impacting the accuracy and representativeness of the results.As mentioned by @groves2012nonresponse, when certain groups are underrepresented due to non-response, it can skew findings, making them less reflective of the actual population.For example, individuals who with demanding schedules, lower income or limited access to technology might be less likely to respond to a poll. Meanwhile, people with strong opinions or from specific demographic groups could be overrepresented in the results. This imbalance can create biased outcomes, where the poll data disproportionately reflects the views of a more vocal subset rather than an accurate cross-section of the entire population.\

@@

Other approaches that SurveyUSA could consider include imputation, as discussed in paper published by @bethlehem2010selection. This method estimates missing responses based on available data, helping to fill gaps and reduce bias in cases of partial non-response.

@@

* **Positive Aspects:**
A logical flow between questions facilitates easy navigation for respondents throughout the survey, while simple wording promotes inclusivity by enabling individuals from diverse backgrounds to comprehend the questions. Furthermore, all questions are directly relevant to analyzing the 2024 U.S. presidential election, and providing predefined response options simplifies the choices for participants.\
* **Negative Aspects:**
Static options for party affiliation and ideology may fail to capture the nuances of respondents’ political beliefs. These rigid categories could oversimplify complex political identities.\

@@

## Objective and Overview
The primary goal of this survey methodology is to forecast the outcome of the U.S. presidential election by collecting high-quality, representative data from a diverse cross-section of American voters. Given a budget of $100,000, this methodology leverages sophisticated sampling techniques, effective respondent recruitment strategies, and robust data validation protocols to ensure maximum accuracy and minimal bias. This approach is designed to account for demographic, geographic, and political factors that can influence voting behavior, thereby enhancing the reliability of predictions [@fivethirty; @pewresearch].

## Core Objectives
* Obtain a representative sample of the U.S. electorate, with demographic diversity mirroring national voting patterns.
* Ensure data quality through rigorous validation processes, including screening and cross-verification.
* Utilize statistical modeling and poll aggregation techniques to provide accurate predictions, as outlined in prior studies of electoral polling

## Sampling Strategy
Our sampling strategy combines **stratified random sampling** with **quota sampling** to ensure broad representation across key demographics. This method has been shown to improve the accuracy of electoral predictions by reducing demographic and geographic sampling bias [@fivethirty].

@@

## Recruitment Strategy
Recruitment is executed through  **multi-channel outreach**, combining digital advertisements, email outreach, and partnerships with civic organizations. This diverse recruitment strategy helps to capture a representative sample and has been identified as a best practice in survey methodologies [@pewresearch].

@@

## Data Validation and Quality Assurance
Maintaining data integrity and ensuring high-quality responses are critical to the accuracy of the election forecast. Several measures will be put in place to validate responses and reduce noise in the dataset.

@@

These practices align with standards for data integrity and quality control in electoral polling [@fivethirty; @pew_survey_methodology].

## Poll Aggregation and Data Analysis
### Poll Aggregation
This survey will be combined with results from reputable polling firms (e.g., YouGov, Ipsos, Gallup) to strengthen the forecast through a **poll-of-polls** approach [@pew_survey_methodology].

@@

Thank you for taking part in this survey aimed at predicting the outcome of the 2024 US Presidential election. Your insights are valuable to our research.

@@

Thanks to Open AI and ChatGPT 4.0 is used to write the paper.
This project utilized @citeR and a variety of R packages that provided essential tools and functionality. We thank the @tidyverse, @ggplot2 and @knitr teams for their extensive suite of tools for data manipulation and visualization, which were instrumental in data cleaning and graphing. We also acknowledge @usmap for its support in creating U.S. geographic visualizations, and @corrplot for facilitating clear, informative correlation plots. Finally, @arrow was crucial for efficient data storage and handling with Parquet files. We are grateful to the developers and maintainers of these packages for their contributions to open-source software.

>>>>

The 2024 U.S. Presidential Election will occur in a context of intensified political polarization, economic challenges, and demographic shifts, positioning it as a significant event in shaping America's future. Accurately forecasting this election’s outcome is essential for understanding potential shifts in political leadership and policy direction. Polling data has traditionally been a primary tool for measuring voter sentiment; however, recent election cycles have revealed limitations in polling accuracy due to biases, non-response, and variations in poll quality. This study proposes a model that addresses these issues to improve the reliability of election forecasts.

This paper develops a predictive model to forecast the electoral outcomes between Kamala Harris and Donald Trump in the 2024 election. Using a multiple linear regression framework, we analyze aggregated polling data across U.S. states, incorporating factors such as poll quality, transparency, sample size, geographic indicators, and poll timing. By integrating these variables, the model aims to produce a detailed forecast that accounts for regional differences and methodological variations, offering a thorough view of projected support for each candidate.

The findings indicate a competitive election landscape, with Harris showing strong prospects in key battleground states, while Trump retains support in traditional Republican strongholds. These projections highlight the persistent influence of geographic voting patterns and the importance of polling methodology in electoral forecasting. This study offers valuable information for political analysts, campaign strategists, and the public, enhancing the understanding of how polling data can inform electoral outcome predictions.

The remainder of this paper is organized as follows: @sec-data describes the dataset and the preprocessing steps for model preparation. @sec-model explains the modeling approach, including predictor selection and the regression model structure. @sec-results presents the primary findings, detailing state-by-state predictions and overall electoral projections. @sec-discussion discusses the implications, limitations, and potential directions for future research. The Appendix provides supplementary information on methodology, diagnostics, technical details of the model setup, acknowledgments, and references.

@@

The dataset used in this analysis originates from @fivethirty, a trusted source known for its rigorous methodology in collecting and aggregating polling data to represent public opinion accurately. To maintain data integrity and ensure a high standard of reliability, only polls meeting stringent quality criteria are included in the dataset. Specifically, each poll must disclose important information, including the pollster’s name, the exact dates of the survey, the sample size, and the full methodological details. These details encompass the polling medium (e.g., phone, online), use of verified voter files, demographic weighting criteria, and any adjustments applied to reflect a representative sample.

@@

Once verified as credible, polls are incorporated into the FiveThirtyEight database. These polls then contribute to polling averages, forecast models, and political analysis, all of which rely on high-quality, reliable public opinion data. This careful selection process ensures that the dataset reflects a thorough and detailed view of public sentiment, accurately capturing behavioral trends and preferences over time.

@@

The distribution shows that most observations cluster around a support level of approximately 48%, indicating significant backing from the electorate. A smaller portion of polls report support above 55%, suggesting that while Trump has a core base, many voters remain either undecided or opposed. This visualization highlights both the concentration and variability of support within the surveyed population.

@@

The sample size represents the number of respondents in each poll. The distribution in @fig-sample-size is right-skewed, indicating a higher frequency of polls with smaller sample sizes. The peak occurs around 1,000 respondents, marking this as the most common sample size among the polls. Overall, the data suggests that each poll includes enough respondents to yield reliable information.

@@

As shown in the @fig-end-date, the support percentage for Donald Trump fluctuated over time, with a slight upward trend as the election approached, indicated by the dashed trend line. This suggests that public opinion varied throughout the period, and Trump’s support percentage experienced an increase as election day drew closer. However, it still remained below 50%. This trend highlights the importance of monitoring real-time polling data, as sentiment may shift in these months leading up to a closely contested election.

@@

The table below (@tbl-prediction) summarizes the predicted results, showing Trump’s predicted percentage, the number of electoral votes in each state, and the predicted winner (either Trump or Harris). For instance:
- Alabama: Trump is predicted to win 53% of the vote, securing all 9 electoral votes.
- California: Trump is predicted to receive 34.39% of the vote, resulting in a victory for Harris, who takes California's 55 electoral votes.
- Florida: The model predicts a close race, with Trump at 48.89% of the vote, resulting in a Harris win in this battleground state.

@@

## Predicted Electoral Outcomes by State
The table in the [Appendix -@tbl-missing-states] supplements the predicted electoral outcomes by providing likely results for states where the model could not generate predictions. Using historical data to estimate outcomes for these missing states ensures a complete view of electoral projections and accounts for every state in the final electoral vote totals.

@@

The analysis shows that state-specific factors, such as historical voting patterns and days until election day, significantly influence each candidate’s polling outcomes. Traditional Democratic strongholds like California and New York favor Harris, while Republican-leaning states such as Alabama and Wyoming align with Trump. This distribution reflects persistent geographic voting patterns and highlights the impact of regional party loyalty.

@@

The model also highlights the significant role of battleground states, with states such as Florida and Arizona predicted to lean toward Harris. The influence of these states, where margins are often narrow, emphasizes the importance of campaign efforts and voter mobilization in these regions. The projections illustrate that minor shifts in voter sentiment within these states could decisively impact the overall election outcome because of the high number of electoral votes.

@@

With limited data for predicting outcomes in every state, some projections are based on historical voting patterns, assuming these trends will continue. However, this approach may be affected by the changing political landscape in the U.S., where evolving demographics and shifting public sentiment could lead to unexpected results.

@@

Non-response in survey research is important to address because it introduces potential bias, impacting the accuracy and representativeness of the results.As mentioned by @groves2012nonresponse, when certain groups are underrepresented due to non-response, it can skew findings, making them less reflective of the actual population.For example, individuals who with demanding schedules, lower income or limited access to technology might be less likely to respond to a poll. Meanwhile, people with strong opinions or from specific demographic groups could be overrepresented in the results. This imbalance can create biased outcomes, where the poll data disproportionately reflects the views of a more vocal subset rather than an accurate cross-section of the entire population.\

@@

Other approaches that SurveyUSA could consider include imputation, as discussed in the work by @bethlehem2010selection. This method estimates missing responses based on existing data, helping to address gaps and reduce bias in cases of partial non-response.

@@

* **Positive Aspects:**
A logical flow between questions facilitates easy navigation for respondents throughout the survey, while simple wording promotes inclusivity by enabling individuals from diverse backgrounds to comprehend the questions. Furthermore, all questions are directly relevant to analyzing the 2024 U.S. presidential election, and providing predefined response options simplifies the choices for participants.\
* **Negative Aspects:**
Static options for party affiliation and ideology may overlook the subtleties of respondents' political beliefs. These fixed categories risk oversimplifying complex political identities.\

@@

## Objective and Overview
The primary goal of this survey methodology is to forecast the U.S. presidential election outcome by collecting representative data from a diverse cross-section of American voters. With a $100,000 budget, this methodology employs sampling techniques, effective recruitment strategies, and data validation protocols to ensure accuracy and minimize bias. This approach is designed to account for demographic, geographic, and political factors that can influence voting behavior, thereby enhancing the reliability of predictions [@fivethirty; @pewresearch].

## Core Objectives
* Obtain a representative sample of the U.S. electorate, with demographic diversity mirroring national voting patterns.
* Ensure data quality through validation processes, including screening and cross-verification.
* Utilize statistical modeling and poll aggregation techniques to provide accurate predictions, as outlined in prior studies of electoral polling

## Sampling Strategy
Our sampling strategy combines **stratified random sampling** with **quota sampling** to ensure broad representation across key demographics. This method improves electoral prediction accuracy by reducing demographic and geographic sampling bias [@fivethirty].

@@

## Recruitment Strategy
Recruitment is executed through  **multi-channel outreach**, combining digital advertisements, email outreach, and partnerships with civic organizations. This recruitment strategy helps capture a representative sample and aligns with best practices in survey methodologies [@pewresearch].

@@

## Data Validation and Quality Assurance
Maintaining data integrity and ensuring high-quality responses are important to the accuracy of the election forecast. Several measures will be put in place to validate responses and reduce noise in the dataset.

@@

These practices align with data integrity and quality control standards in electoral polling [@fivethirty; @pew_survey_methodology].

## Poll Aggregation and Data Analysis
### Poll Aggregation
This survey will be combined with results from reputable polling firms (e.g., YouGov, Ipsos, Gallup) using a **poll-of-polls** approach to strengthen the forecast [@pew_survey_methodology].

@@

Thank you for taking part in this survey aimed at predicting the outcome of the 2024 US Presidential election. Your voices are valuable to our research.

@@

Thanks to Open AI and ChatGPT 4.0 is used to write the paper.
This project utilized @citeR and a variety of R packages that provided essential tools and functionality. We thank the @tidyverse, @ggplot2 and @knitr teams for their extensive suite of tools for data manipulation and visualization, which were instrumental in data cleaning and graphing. We also acknowledge @usmap for its support in creating U.S. geographic visualizations, and @corrplot for facilitating clear, informative correlation plots. Finally, @arrow was vital for efficient data storage and handling with Parquet files. We are grateful to the developers and maintainers of these packages for their contributions to open-source software.
